---
title: "Workflow_WTA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow_WTA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(GeoDiff)
library(dplyr)
library(pheatmap)
library(ggplot2)
library(NanoStringNCTools)
library(GeomxTools)
library(Biobase)
```



# Data

```{r}
data("kidney")
# alter featureNames
stopifnot(identical(rownames(fData(kidney)), featureNames(kidney)))
stopifnot(identical(rownames(fData(kidney)), rownames(exprs(kidney))))

rownames(fData(kidney))[which(!fData(kidney)$Negative)] <- fData(kidney)[which(!fData(kidney)$Negative), "TargetName"]
featureNames(kidney) <- rownames(fData(kidney))
rownames(exprs(kidney)) <- rownames(fData(kidney))
```


In order to make the vignette complete running in a reasonable amount of time, we subset the data. 
```{r}
kidney <- kidney[, which(kidney$`slide name` %in% c("disease3", "normal3"))][, c(1:4, 48:51, 60:63, 115:118)]
table(kidney$region, kidney$`slide name`)
table(kidney$`slide name`, kidney$class)
```


## Poisson Background model for negative probes

```{r}
kidney <- fitPoisBG(kidney, size_scale = "sum")
```

```{r}
set.seed(123)
kidney_diag <- diagPoisBG(kidney, split = FALSE)

notes(kidney_diag)$disper
```

```{r}
which(assayDataElement(kidney_diag, "low_outlier") == 1, arr.ind = TRUE)
which(assayDataElement(kidney_diag, "up_outlier") == 1, arr.ind = TRUE)
```

```{r}
kidney <- fitPoisBG(kidney, groupvar = "slide name", size_scale = "sum")
```

```{r}
set.seed(123)
kidney_diag <- diagPoisBG(kidney, split = TRUE)
notes(kidney_diag)$disper_sp
```

```{r}
which(assayDataElement(kidney_diag, "low_outlier") == 1, arr.ind = TRUE)
which(assayDataElement(kidney_diag, "up_outlier") == 1, arr.ind = TRUE)
```


# Score test


`scoretest_sp` can't deal with outlier detection at this moment.

```{r}

kidney <- BGScoreTest(kidney, split = FALSE, adj = 1, removeoutlier = FALSE, useprior = FALSE)

kidney1o <- BGScoreTest(kidney, split = FALSE, adj = 1, removeoutlier = TRUE, useprior = FALSE)
kidney1p <- BGScoreTest(kidney, split = FALSE, adj = 1, removeoutlier = FALSE, useprior = TRUE)
kidney1op <- BGScoreTest(kidney, split = FALSE, adj = 1, removeoutlier = TRUE, useprior = TRUE)
kidney2 <- BGScoreTest(kidney, split = TRUE, adj = 1, removeoutlier = FALSE, useprior = FALSE)
kidney2p <- BGScoreTest(kidney, split = TRUE, adj = 1, removeoutlier = FALSE, useprior = TRUE)
kidney2o <- BGScoreTest(kidney, split = TRUE, adj = 1, removeoutlier = TRUE, useprior = FALSE)
kidney2op <- BGScoreTest(kidney, split = TRUE, adj = 1, removeoutlier = TRUE, useprior = TRUE)
```


## Number of genes above the background by each method

Notice: high percentage of genes considered above the background by score test in this dataset. Also the score test with prior is more conservative but also prone to be influenced by outliers.

```{r}
sum(fData(kidney)[["pvalues"]] < 1e-3, na.rm = TRUE)
mean(fData(kidney)[["pvalues"]] < 1e-3, na.rm = TRUE)

sum(fData(kidney1o)[["pvalues"]] < 1e-3, na.rm = TRUE)
mean(fData(kidney1o)[["pvalues"]] < 1e-3, na.rm = TRUE)

sum(fData(kidney1p)[["pvalues"]] < 1e-3, na.rm = TRUE)
mean(fData(kidney1p)[["pvalues"]] < 1e-3, na.rm = TRUE)

sum(fData(kidney1op)[["pvalues"]] < 1e-3, na.rm = TRUE)
mean(fData(kidney1op)[["pvalues"]] < 1e-3, na.rm = TRUE)

colSums(fData(kidney2)[, grep("pvalues_", fvarLabels(kidney2))] < 1e-3, na.rm = TRUE)
colMeans(fData(kidney2)[, grep("pvalues_", fvarLabels(kidney2))] < 1e-3, na.rm = TRUE)

colSums(fData(kidney2o)[, grep("pvalues_", fvarLabels(kidney2o))] < 1e-3, na.rm = TRUE)
colMeans(fData(kidney2o)[, grep("pvalues_", fvarLabels(kidney2o))] < 1e-3, na.rm = TRUE)

colSums(fData(kidney2p)[, grep("pvalues_", fvarLabels(kidney2p))] < 1e-3, na.rm = TRUE)
colMeans(fData(kidney2p)[, grep("pvalues_", fvarLabels(kidney2p))] < 1e-3, na.rm = TRUE)

colSums(fData(kidney2op)[, grep("pvalues_", fvarLabels(kidney2op))] < 1e-3, na.rm = TRUE)
colMeans(fData(kidney2op)[, grep("pvalues_", fvarLabels(kidney2op))] < 1e-3, na.rm = TRUE)

```


# Estimate the scaling factor

To estimate the scaling factor, we need to define a set of highly expressed genes. One way to do that is to use score test defined high expressed genes. However in this dataset the score test gives too many features_high so we will just select high genes using quantiles of total count of genes.

```{r, cache=TRUE}
set.seed(123)

kidney <- fitNBth(kidney,
  iterations = 5,
  start_para = c(200, 1),
  lower_sizefact = 0,
  lower_threshold = 100, tol = 1e-8
)
```

There are three arguments (features_high, sizefact_BG, threshold_start) in the fitNBth function that can be modified if needed. Below is the example in which we provide specific values rather than using the default values. 

```{r}
posdat <- kidney[-which(fData(kidney)$CodeClass == "Negative"), ]
posdat <- exprs(posdat)
gene_sum <- rowSums(posdat)

kidney_neg <- kidney[which(fData(kidney)$CodeClass == "Negative"), ]
featfact <- fData(kidney_neg)[, "featfact"]

features_high <- ((gene_sum>quantile(gene_sum, probs = 0.5)) & (gene_sum<quantile(gene_sum, probs = 0.95))) %>% which %>% names

set.seed(123)

features_high <- sample(features_high, 1500)
```

We also need the mean of the background, since all gene has 1 probes so adjustment factor=1.

```{r}
thmean <- 1*mean(featfact)
```

Fit one scaling factor for all slides.

```{r, cache=TRUE}
set.seed(123)

kidney <- fitNBth(kidney, 
                  features_high=features_high, 
                  sizefact_BG = kidney$sizefact, 
                  threshold_start = thmean, 
                  iterations=5, 
                  start_para = c(200, 1), 
                  lower_sizefact=0, 
                  lower_threshold=100, tol=1e-8,
                  threshold_fix=FALSE)

kidney2 <- fitNBth(kidney, 
                  features_high=features_high, 
                  sizefact_BG = kidney$sizefact, 
                  threshold_start = thmean, 
                  iterations=5, 
                  start_para = c(200, 1), 
                  lower_sizefact=0, 
                  lower_threshold=100, tol=1e-8,
                  threshold_fix=TRUE)
```

```{r}
notes(kidney)[["threshold"]]
thmean
```

```{r}
plot(kidney$sizefact, kidney2$sizefact)
abline(a=0,b=1)

```

```{r}
plot(fData(kidney)[["para"]][features_high,1], 1/fData(kidney)[["para"]][features_high,2])
lines(lowess(1/fData(kidney)[["para"]][features_high,2] ~ fData(kidney)[["para"]][features_high,1], delta=2), col=2, lty=2)

```


In this dataset, this size factor correlate well with $90\%$ quantile.

```{r}
cor(kidney$sizefact, kidney$sizefact_fitNBth)
plot(kidney$sizefact, kidney$sizefact_fitNBth)
abline(a = 0, b = 1)
quan <- sapply(c(0.75, 0.8, 0.9, 0.95), function(y) apply(posdat, 2, function(x) quantile(x, probs = y)))

apply(quan, 2, function(x) cor(x, kidney$sizefact_fitNBth))

quan75 <- apply(posdat, 2, function(x) quantile(x, probs = 0.75))
```

Quantile range has even better correlation with the size factor


```{r}
kidney <- QuanRange(kidney, split = FALSE, probs = c(0.75, 0.8, 0.9, 0.95))
```

```{r}
apply(pData(kidney)[, as.character(c(0.75, 0.8, 0.9, 0.95))], 2, function(x) cor(x, kidney$sizefact_fitNBth))
```

```{r}
ROIs_high <- sampleNames(kidney)[which((quantile(fData(kidney)[["para"]][, 1],
                                                  probs = 0.90, na.rm = TRUE) - notes(kidney)[["threshold"]])*kidney$sizefact_fitNBth>2)]
 
features_all <- rownames(posdat)

min(fData(kidney)[["para"]][ , 1], na.rm = TRUE)

```


# DE model

Previous recommendation. The problem of it is that this model specification is not parameterization invariant, that is if you scale alpha and beta differently in Poisson model for negative probes, among the estimated parameters in the DEmod, the intercept will be shifted every time. This is really undesirable and makes interpretation of the parameters really difficult. While this model call is kept here for legacy purpose, we do not recommend call this function this way anymore.


```{r, cache=TRUE}

NBthDEmod1 <- fitNBthDE(form = ~region, 
                        split = FALSE, 
                        object = kidney,
                        ROIs_high = ROIs_high, 
                        features_high = features_high, 
                        features_all = features_all,
                        sizefact_start = kidney[, ROIs_high][['sizefact_fitNBth']],
                        sizefact_BG = kidney[, ROIs_high][['sizefact']], 
                        threshold_mean = 1, 
                        preci2=10000, 
                        prior_type="contrast", 
                        covrob=FALSE, 
                        preci1con=1/(5+log2(notes(kidney)[["threshold"]]))^2,
                        sizescalebythreshold = FALSE)

```

Current recommendation, in which both size factors are multiplied by thmean, which means that no matter how you parameterize size factor and thmean in Poisson model for negative probes, you end up with the same parameter estimation. 

```{r, cache=TRUE}

NBthDEmod2 <- fitNBthDE(form = ~region, 
                        split = FALSE, 
                        object = kidney,
                        ROIs_high = ROIs_high, 
                        features_high = features_high, 
                        features_all = features_all,
                        sizefact_start = kidney[, ROIs_high][['sizefact_fitNBth']],
                        sizefact_BG = kidney[, ROIs_high][['sizefact']], 
                        threshold_mean = notes(kidney)[["threshold"]], 
                        preci2=10000, 
                        prior_type="contrast", 
                        covrob=FALSE, 
                        preci1con=1/25, 
                        sizescalebythreshold=TRUE)

```


```{r, cache=TRUE}
NBthDEmod3 <- fitNBthDE(form = ~region, 
                        split = FALSE, 
                        object = kidney,
                        features_high = features_high, 
                        features_all = features_all,
                        sizefact_start = kidney[, ROIs_high][['sizefact_fitNBth']],
                        sizefact_BG = kidney[, ROIs_high][['sizefact']], 
                        threshold_mean = notes(kidney)[["threshold"]], 
                        preci2=10000, 
                        prior_type="contrast", 
                        covrob=FALSE, 
                        preci1con=10, 
                        sizescale=TRUE)

```


Relationship between two DE model fit
```{r}
plot(NBthDEmod1$para[1,], NBthDEmod2$para[1,])
abline(a=-log2(notes(kidney)[["threshold"]]),b=1)
#The log2 FC with respect of the region variable
plot(NBthDEmod1$para[2,], NBthDEmod2$para[2,])
abline(a=0,b=1)
```

```{r}
plot(NBthDEmod2$para[2,], NBthDEmod3$para[2,])
abline(a=0,b=1)
```

```{r}
sc1_scores <- fData(kidney)[-which(fData(kidney)[["Negative"]]), "scores"]
names(sc1_scores) <- fData(kidney)[-which(fData(kidney)[["Negative"]]), "TargetName"]
```


```{r}
plot(sc1_scores, NBthDEmod2$para[2,], xlim=c(-50,200))
```

```{r}
plot(sc1_scores, NBthDEmod3$para[2,], xlim=c(-50,200))
```


# Mixed effect model

First take a look at the study design. It shows the two levels of region both exist in the same patient ID. This indicate the random effect model with random slope would be appropriate, still we fit both random intercept model and random slope model to showcase the capability of the mixed model function. Both mixed model function calls take quite some time so the model objects are saved and can be directly loaded to inspection.

```{r}
pData(kidney)$region <- factor(pData(kidney)$region, levels=c("glomerulus", "tubule"))

table(pData(kidney)[, c("region", "slide name")])

features_high_subset <- features_high[1:30]
```

Random intercept model only for high genes as an example, takes about 1 hour. Users can apply this function to the whole gene set but it takes even longer.


```{r}
set.seed(123)
NBthmDEmod2 <- fitNBthmDE(object = kidney,
                          form = ~ region+(1|`slide name`), 
                          ROIs_high = ROIs_high, 
                          split = FALSE,
                          features_all = features_high_subset, 
                          sizefact = kidney[, ROIs_high][['sizefact_fitNBth']],
                          sizefact_BG=kidney[, ROIs_high][['sizefact']],
                          preci1=NBthDEmod2$preci1, 
                          threshold_mean = thmean,
                          preci2=10000, 
                          sizescale = TRUE, 
                          controlRandom=list(nu=12, nmh_e=400, thin_e=60))
```




Random slope model(recommended for this study design), takes about 4 hours. Users can apply this function to the whole gene set but it takes even longer.

Users can call

```{r}
set.seed(123)
NBthmDEmod2slope <- fitNBthmDE(object = kidney, 
                               form = ~ region+(1+region|`slide name`), 
                               split = FALSE,
                               ROIs_high = ROIs_high, 
                               features_all = features_high_subset, 
                               sizefact = kidney[, ROIs_high][['sizefact_fitNBth']],
                               sizefact_BG=kidney[, ROIs_high][['sizefact']],
                               preci1=NBthDEmod2$preci1, 
                               threshold_mean = thmean, 
                               preci2=10000, 
                               sizescale = TRUE, 
                               controlRandom=list(nu=12, nmh_e=400, thin_e=60))

```



Relation between random effect and fixed effect models.

```{r}
plot(NBthDEmod2$para[2,names(NBthmDEmod2$para[2,])], NBthmDEmod2$para[2,])
abline(a=0,b=1)

plot(NBthDEmod2$para[2,names(NBthmDEmod2slope$para[2,])], NBthmDEmod2slope$para[2,])
abline(a=0,b=1)

```

Genes with larger difference in estimates between fixed effect model and random slope model have larger random effect variance for the random slope. 


```{r}
diff_high <- names(which(abs(NBthDEmod2$para[2,names(NBthmDEmod2slope$para[2,])]-NBthmDEmod2slope$para[2,])>1.5))
diff_high
set.seed(123)

```


```{r}
NBthmDEmod2slope$theta[3, diff_high]
```


```{r}
annot <- pData(kidney)
annot$KIRREL1 <- posdat["KIRREL1",]
annot$KIRREL1 <- posdat["KIRREL1",]

```

```{r}
NBthDEmod2$para[2,"KIRREL1"]
NBthmDEmod2$para[2,"KIRREL1"]
NBthmDEmod2slope$para[2,"KIRREL1"]
```

The figure below shows there are huge variation in the difference between two levels of region within each slide.

```{r}

NBthmDEmod2slope$theta[,"KIRREL1"]
plot_dat <- annot[,c("region", "KIRREL1", "slide name")]
p <- ggplot(plot_dat, aes(x=`slide name`, y=KIRREL1, fill=region)) + 
  geom_boxplot()

plot(p)



```







# Generate DE result

DE tables


```{r}
coeff <- coefNBth(NBthDEmod2)
coeffhighpen <- coefNBth(NBthDEmod3)
coefr <- coefNBth(NBthmDEmod2)
coefrslope <- coefNBth(NBthmDEmod2slope)

```



```{r}
is.na(coeff$p_value)%>%sum
is.na(coeffhighpen$p_value)%>%sum
is.na(coefr$p_value)%>%sum
is.na(coefrslope$p_value)%>%sum
```


log2 FC and p values for region

```{r}


DEtab <- DENBth(coeff, variable = "regiontubule")
DEtabhighpen <- DENBth(coeffhighpen, variable = "regiontubule")
DEtabr <- DENBth(coefr, variable = "regiontubule")
DEtabrslope <- DENBth(coefrslope, variable = "regiontubule")

```

Relation between p values and scores


```{r}

plot(sc1_scores, DEtab$pvalue, xlim=c(min(sc1_scores),100))
plot(sc1_scores, DEtab$adjp, xlim=c(min(sc1_scores),100))

plot(sc1_scores, DEtabhighpen$pvalue, xlim=c(min(sc1_scores),100))
plot(sc1_scores, DEtabhighpen$adjp, xlim=c(min(sc1_scores),100))

plot(sc1_scores[features_high_subset], DEtabr$pvalue, xlim=c(min(sc1_scores),100))
plot(sc1_scores[features_high_subset], DEtabr$adjp, xlim=c(min(sc1_scores),100))

plot(sc1_scores[features_high_subset], DEtabrslope$pvalue, xlim=c(min(sc1_scores),100))
plot(sc1_scores[features_high_subset], DEtabrslope$adjp, xlim=c(min(sc1_scores),100))

```



Relation between p values from different models. The random slope model less significant p values than the other two.

```{r}
plot(log10(DEtab[,"pvalue"]), log10(DEtabhighpen[,"pvalue"]), xlim=c(-100,0), ylim=c(-100,0))
abline(a=0,b=1)

plot(log10(DEtab[features_high_subset,"pvalue"]), log10(DEtabr[,"pvalue"]), xlim=c(-100,0), ylim=c(-100,0))

plot(log10(DEtab[features_high_subset,"pvalue"]), log10(DEtabrslope[,"pvalue"]),xlim=c(-100,0), ylim=c(-100,0))


plot(log10(DEtabr[,"pvalue"]), log10(DEtabrslope[,"pvalue"]),xlim=c(-100,0), ylim=c(-100,0))
```



# Normalization

Not consider different slides

```{r}
set.seed(123)

kidney <- fitPoisthNorm(
  object = kidney,
  threshold_mean = 1,
  preci2 = 10000,
  prior_type = "contrast",
  covrob = FALSE,
  preci1con = 1 / 25
)
```

There are five arguments (ROIs_high, features_high, features_all, sizefact_start, sizefact_BG) in the fitPoisthNorm function that can be modified if needed. Below is the example in which we provide specific values rather than using the default values. 

```{r, cache=TRUE}

features_all <- rownames(posdat)

```

```{r}
set.seed(123)

kidney <- fitPoisthNorm(object = kidney, 
                        split = FALSE, 
                        ROIs_high = ROIs_high, 
                        features_high = features_high, 
                        features_all = features_all,
                        sizefact_start = kidney[, ROIs_high][['sizefact_fitNBth']],
                        sizefact_BG = kidney[, ROIs_high][['sizefact']], 
                        threshold_mean = thmean, 
                        preci2=10000, 
                        prior_type = "contrast", 
                        covrob=FALSE, 
                        preci1con=1/25,
                        sizescalebythreshold=TRUE)


```

Correlation between normalized data without and with EB prior.

```{r}
N <- length(ROIs_high)

corvec0 <- apply(cbind(
  assayDataElement(kidney, "normmat0")[features_high, ROIs_high],
  assayDataElement(kidney, "normmat")[features_all, ROIs_high][features_high, ]
), 1, function(x) cor(x[1:N], x[(N+1):(2*N)]))
```


Score vs correlation, shows the EB prior has marginal impact on the normalized expression for high genes.

```{r}
plot(fData(kidney)[features_high, "scores"], corvec0, ylim = c(-1, 1))
```



Normalize the dataset within each batch/slide.

```{r, cache=TRUE}
set.seed(123)

kidney <- fitPoisthNorm(
                object = kidney, 
                split = TRUE, 
                ROIs_high = ROIs_high, 
                features_high = features_high, 
                features_all = features_all,
                sizefact_start = kidney[, ROIs_high][['sizefact_fitNBth']],
                sizefact_BG = kidney[, ROIs_high][['sizefact']], 
                threshold_mean = thmean, 
                preci2=10000, 
                prior_type="contrast", 
                covrob=FALSE, 
                preci1con=1/25, 
                sizescalebythreshold=TRUE
                )

```





Compared to background subtraction and Q75 normalization


```{r}

norm_dat_backqu75 <- sweep(posdat[, ROIs_high], 2, (kidney[, ROIs_high]$sizefact * thmean), FUN = "-") %>%
  sweep(., 2, quan75[ROIs_high], FUN = "/") %>%
  pmax(., 0) %>%
  `+`(., 0.01) %>%
  log2()
```


```{r}

dat_plot <- cbind(pData(kidney)[ROIs_high, c("slide name", "region")],
                  t(norm_dat_backqu75[features_all, ]))

dat_plot <- cbind(dat_plot, ROI_ID = ROIs_high)

library(reshape2)

# dat_plot$Resection <- factor(dat_plot$Resection, levels=c("Control", "Primary", "Recurrent"))


dat_plot <- melt(dat_plot, id.vars = c("ROI_ID", "slide name", "region"))
# dat_plot <- dat_plot[order(dat_plot$Scan.name, dat_plot$Resection),]




ggplot(dat_plot, aes(x = value)) +
  geom_density(aes(fill = region, group = ROI_ID, alpha = 0.01)) +
  facet_wrap(~`slide name`) +
  ggtitle("backqu75")
```

```{r}
dat_plot <- cbind(annot[ROIs_high, c("slide name", "region")], 
                  t(assayDataElement(kidney[features_high, ROIs_high], "normmat_sp")))

dat_plot <- cbind(dat_plot, ROI_ID = ROIs_high)



# dat_plot$Resection <- factor(dat_plot$Resection, levels=c("Control", "Primary", "Recurrent"))


dat_plot <- melt(dat_plot, id.vars = c("ROI_ID", "slide name", "region"))
# dat_plot <- dat_plot[order(dat_plot$Scan.name, dat_plot$Resection),]




ggplot(dat_plot, aes(x = value)) +
  geom_density(aes(fill = region, group = ROI_ID, alpha = 0.01)) +
  facet_wrap(~`slide name`) +
  ggtitle("Poisson threshold normalization")
```
```{r}
kidney <- BGScoreTest(kidney, split = TRUE, adj = 1, removeoutlier = FALSE, useprior = TRUE)

features_sig <- apply(fData(kidney)[, grep("pvalues_", fvarLabels(kidney))] < 0.001, 1, all)

features_sig <- names(which(features_sig))
```


```{r}
dat_plot <- cbind(annot[ROIs_high, c("slide name", "region")], t(norm_dat_backqu75[features_sig, ]))

dat_plot <- cbind(dat_plot, ROI_ID = ROIs_high)

library(reshape2)

# dat_plot$Resection <- factor(dat_plot$Resection, levels=c("Control", "Primary", "Recurrent"))


dat_plot <- melt(dat_plot, id.vars = c("ROI_ID", "slide name", "region"))
# dat_plot <- dat_plot[order(dat_plot$Scan.name, dat_plot$Resection),]




ggplot(dat_plot, aes(x = value)) +
  geom_density(aes(fill = region, group = ROI_ID, alpha = 0.01)) +
  facet_wrap(~`slide name`) +
  ggtitle("backqu75")
```

```{r}
dat_plot <- cbind(annot[ROIs_high, c("slide name", "region")], 
                  t(assayDataElement(kidney[features_sig, ROIs_high], "normmat_sp")))

dat_plot <- cbind(dat_plot, ROI_ID = ROIs_high)



# dat_plot$Resection <- factor(dat_plot$Resection, levels=c("Control", "Primary", "Recurrent"))


dat_plot <- melt(dat_plot, id.vars = c("ROI_ID", "slide name", "region"))
# dat_plot <- dat_plot[order(dat_plot$Scan.name, dat_plot$Resection),]




ggplot(dat_plot, aes(x = value)) +
  geom_density(aes(fill = region, group = ROI_ID, alpha = 0.01)) +
  facet_wrap(~`slide name`) +
  ggtitle("Poisson threshold normalization")
```




```{r}
dat <- t(norm_dat_backqu75[features_high, ])
dat_pca <- prcomp(dat, center = TRUE, scale. = TRUE)
dat <- as.data.frame(dat)

dat$PC1 <- dat_pca$x[, 1]
dat$PC2 <- dat_pca$x[, 2]
dat$id <- annot$`slide name`[match(ROIs_high, colnames(posdat))]
dat$class <- annot$class[match(ROIs_high, colnames(posdat))]
dat$region <- annot$region[match(ROIs_high, colnames(posdat))]
dat$sizeratio <- kidney[, ROIs_high]$sizefact_fitNBth / kidney[, ROIs_high]$sizefact

p <- ggplot(data = dat, aes(x = PC1, y = PC2)) +
  geom_point(aes(colour = class)) +
  theme_bw()

plot(p)

p <- ggplot(data = dat, aes(x = PC1, y = PC2)) +
  geom_point(aes(colour = region)) +
  theme_bw()

plot(p)


p <- ggplot(data = dat, aes(x = PC1, y = PC2)) +
  geom_point(aes(colour = log2(sizeratio))) +
  theme_bw()

plot(p)
```


```{r}
dat <- t(assayDataElement(kidney[features_high, ROIs_high],"normmat_sp"))
dat_pca <- prcomp(dat, center = TRUE, scale. = TRUE)
dat <- as.data.frame(dat)

dat$PC1 <- dat_pca$x[, 1]
dat$PC2 <- dat_pca$x[, 2]
dat$id <- annot$`slide name`[match(ROIs_high, colnames(posdat))]
dat$class <- annot$class[match(ROIs_high, colnames(posdat))]
dat$region <- annot$region[match(ROIs_high, colnames(posdat))]
dat$sizeratio <- kidney[, ROIs_high]$sizefact_fitNBt / kidney[, ROIs_high]$sizefact

p <- ggplot(data = dat, aes(x = PC1, y = PC2)) +
  geom_point(aes(colour = class)) +
  theme_bw()

plot(p)

p <- ggplot(data = dat, aes(x = PC1, y = PC2)) +
  geom_point(aes(colour = id)) +
  theme_bw()

plot(p)

p <- ggplot(data = dat, aes(x = PC1, y = PC2)) +
  geom_point(aes(colour = region)) +
  theme_bw()

plot(p)

p <- ggplot(data = dat, aes(x = PC1, y = PC2)) +
  geom_point(aes(colour = log2(sizeratio))) +
  theme_bw()

plot(p)
```


```{r}
sessionInfo()
```

